\section{Circuit Minimization}

\subsection{Intro}
This is the process of representing a Boolean function with the fewest possible
products of literals, where these products contain the fewest literals. This 
makes it possible for the construction of a circuit that uses the minimum 
number of inputs and gates to occur which in the early 1960s computing was 
necessary to reduce costs and run more efficiently.

\bigskip
\noindent\textbf{Sum-of-Products (differing by only one variable)}

\begin{itemize}
    \item $xyz + xy'z = (y+y')xz = xz$
    \item $wxyz' + wxyz = (z'+z)wxy = wxy$
\end{itemize}

\bigskip
\noindent The above examples can be represented by circuits in many ways but by 
minimization of circuits we will represent them by the smallest number of 
gates and inputs as shown later.

\subsection{Complexity of Boolean Minimization}
Minimizing Boolean functions with many variables is a computationally intensive
task as it has been shown to be a NP-complete problem (meaning existence of a 
fast (polynomial time) algorithm is unlikely) and of current algorithms can only
minimize up to 25 variables \cite{shannon1938}.

\subsection{Why Minimization Matters}
Circuit minimization is crucial in digital logic design because it:
\begin{itemize}
    \item Reduces the number of logic gates required
    \item Leads to lower cost of implementation
    \item Results in less power consumption
    \item Produces faster operation due to reduced delay
    \item Makes the overall system more efficient and reliable
\end{itemize}

\subsection{Minimization Techniques}

\begin{enumerate}
    \item Karnaugh Maps (K-Maps)
    \item Quine-McCluskey Method
\end{enumerate}

\section{Karnaugh Maps (K-Maps)}

This is a visual method of simplifying sum-of-products expansions for Boolean 
functions with a smaller number of variables ($\leq 6$) by combining terms. To 
minimize, we identify blocks of 1's in the map and use the corresponding 
variables \cite{stem2024}.

\subsection{Key Notes}

\begin{itemize}
    \item A block must contain $2^k$ cells (usually: 1, 2, 4, 8, 16).
    \item Identify the largest possible blocks and cover all the 1's with the 
    least number of blocks.
    \item \textbf{Implicant:} A product of literals corresponding to a block of 
    all 1's.
    \item \textbf{Prime Implicant:} A block of 1's that is not contained in any 
    larger block of all 1's.
    \item \textbf{Essential Prime Implicant:} A prime implicant that covers at 
    least one cell containing a 1 that is not covered by any other prime 
    implicant and it must be included in the final expression.
    \item \textbf{Literal elimination:} A block of $2^k$ cells in an $n$-variable 
    map results in $n-k$ literals.
\end{itemize}

\subsection{Steps to Use K-Maps}

\begin{enumerate}
    \item \textbf{Choose the K-map:} Select the correct K-map based on the 
    number of variables (e.g., 2, 3, or 4 variables).
    
    \item \textbf{Identify minterms or maxterms:} Determine which combinations of 
    inputs produce a 1 (minterms) or 0 (maxterms) from the given function.
    
    \item \textbf{Fill in the K-map:} For SOP, place 1s in the cells 
    corresponding to the minterms and 0s elsewhere.
    
    \item \textbf{Form groups:} Group adjacent cells containing 1s. Each group 
    must have a size that is a power of 2 (e.g., 2, 4, 8).
    
    \item \textbf{Follow grouping rules:} Groups must be as large as possible, 
    no diagonal grouping is allowed, overlapping is allowed, and wrapping around 
    edges is permitted.
    
    \item \textbf{Derive simplified terms:} For each group, identify the 
    variables that remain constant and eliminate those that change.
    
    \item \textbf{Write the final expression:} Combine all simplified terms to 
    form the minimized Boolean expression.
\end{enumerate}

\subsection{2-Variable K-map}

Given $F(x,y) = xy + x'y$, we can simplify:

\begin{align*}
    xy + x'y &= (y + y')x \\
             &= x
\end{align*}

\begin{center}
\begin{tabular}{c|cc|c}
    \textbf{yy'} & \textbf{x} & \textbf{x'} & \\
    \hline
    \textbf{x}     & 1 & 1 & $x$ \\
    \textbf{x'}    & 0 & 1 & $x'$ \\
    \hline
               & $y$ & $y'$ & 
\end{tabular}
\qquad
\begin{tabular}{c|cc|c}
    \textbf{yy'} & \textbf{x} & \textbf{x'} & \\
    \hline
    \textbf{x}     & 1 & 0 & $xy$ \\
    \textbf{x'}    & 1 & 1 & $x'y + x'y'$ \\
    \hline
               & $y$ & $y'$ & 
\end{tabular}
\end{center}

\bigskip
\noindent Given $F(x,y) = xy' + x'y$:

\begin{align*}
    xy' + x'y + x'y' &= x'(y + y') + xy' \\
                    &= x' + xy'
\end{align*}

\begin{center}
\begin{tabular}{c|cc|c}
    \textbf{yy'} & \textbf{x} & \textbf{x'} & \\
    \hline
    \textbf{x}     & 0 & 1 & $x'$ \\
    \textbf{x'}    & 1 & 1 & $x' + y'$ \\
    \hline
               & $y$ & $y'$ & 
\end{tabular}
\qquad
\begin{tabular}{c|cc|c}
    \textbf{yy'} & \textbf{x} & \textbf{x'} & \\
    \hline
    \textbf{x}     & 1 & 0 & $xy'$ \\
    \textbf{x'}    & 1 & 1 & $x'y + x'y'$ \\
    \hline
               & $y$ & $y'$ & 
\end{tabular}
\end{center}

\subsection{3-Variable K-map}

\bigskip
\noindent\textbf{Example 1:} Simplify $F(x,y,z) = xyz' + xy'z' + x'yz + x'y'z'$

\begin{align*}
    &= xz'(y + y') + x'z(y + y') \\
    &= xz' + x'z
\end{align*}

\begin{center}
\begin{tabular}{c|cccc|c}
    \textbf{yz} & \textbf{00} & \textbf{01} & \textbf{11} & \textbf{10} & \\
    \hline
    \textbf{0} & 1          & 0          & 1          & 0          & $x'$ \\
    \textbf{1} & 1          & 0          & 1          & 1          & $xz' + x'z$ \\
    \hline
               & $z'$       & $z$        &            &            & 
\end{tabular}
\end{center}

\bigskip
\noindent\textbf{Example 2:} Simplify $F(x,y,z) = xy'z + xy'z' + x'yz + x'y'z + x'y'z'$

\begin{align*}
    &= y'(xz + xz') + x'(yz + y'z) \\
    &= y'x(z + z') + x'(y + y')z \\
    &= y' + x'z
\end{align*}

\begin{center}
\begin{tabular}{c|cccc|c}
    \textbf{yz} & \textbf{00} & \textbf{01} & \textbf{11} & \textbf{10} & \\
    \hline
    \textbf{0} & 1          & 1          & 1          & 0          & $y' + x'z$ \\
    \textbf{1} & 0          & 1          & 1          & 1          & $y'$ \\
    \hline
               & $z'$       & $z$        &            &            & 
\end{tabular}
\end{center}

\bigskip
\noindent\textbf{Example 3:} Simplify $F(x,y,z) = xyz + xyz' + xy'z + xy'z' + x'yz + x'y'z + x'y'z'$

\begin{align*}
    &= xy(z + z') + xy'(z + z') + x'y(z + z') + x'y'z \\
    &= xy + xy' + x'y + x'y'z \\
    &= x(y + y') + x'(y + y'z) \\
    &= x + x'y \\
    &= x + y
\end{align*}

\begin{center}
\begin{tabular}{c|cccc|c}
    \textbf{yz} & \textbf{00} & \textbf{01} & \textbf{11} & \textbf{10} & \\
    \hline
    \textbf{0} & 1          & 1          & 1          & 1          & $x + y$ \\
    \textbf{1} & 0          & 1          & 1          & 1          & $x + y$ \\
    \hline
               & $z'$       & $z$        &            &            & 
\end{tabular}
\end{center}

\bigskip
\noindent\textbf{Example 4:} Simplify $F(x,y,z) = xyz' + xy'z' + x'y'z + x'y'z'$

\begin{align*}
    &= xz'(y + y') + x'y'(z + z') \\
    &= xz' + x'y'
\end{align*}

\begin{center}
\begin{tabular}{c|cccc|c}
    \textbf{yz} & \textbf{00} & \textbf{01} & \textbf{11} & \textbf{10} & \\
    \hline
    \textbf{0} & 1          & 0          & 0          & 1          & $x'z' + xy'$ \\
    \textbf{1} & 0          & 0          & 1          & 1          & $xz'$ \\
    \hline
               & $z'$       & $z$        &            &            & 
\end{tabular}
\end{center}

\subsection{4-Variable K-map}

\bigskip
\begin{center}
\begin{tabular}{c|ccccccc|c}
    \textbf{yz} & \textbf{00} & \textbf{01} & \textbf{11} & \textbf{10} & & & \\
    \hline
    \textbf{wx = 00}    & $w'x'y'z'$ & $w'x'y'z$ & $w'x'yz$ & $w'x'yz'$ & & & $w'x'$ \\
    \textbf{wx = 01}    & $w'xy'z'$ & $w'xy'z$ & $w'xyz$ & $w'xyz'$ & & & $w'x$ \\
    \textbf{wx = 11}    & $wxy'z'$ & $wxy'z$ & $wxyz$ & $wxyz'$ & & & $wx$ \\
    \textbf{wx = 10}    & $wx'y'z'$ & $wx'y'z$ & $wx'yz$ & $wx'yz'$ & & & $wx'$ \\
    \hline
                       & $y'z'$    & $y'z$     & $yz$     & $yz'$     & & &
\end{tabular}
\end{center}

\subsection{K-map Notes}

\begin{itemize}
    \item For $n$ variables, there are $2^n$ possible minterms in the map.
    \item We can use gray codes for plotting variables into the map to make sure 
    only one literal changes between two adjacent columns.
    \item We eliminate the variable that changes:
    \begin{itemize}
        \item A block of 2 results in 1 variable eliminated.
        \item A block of 4 results in 2 variables eliminated.
        \item A block of 8 leads to 3 variables eliminated, giving constant 1.
    \end{itemize}
    \item If a minterm is present in a function, the cell gets a 1; if not, it 
    gets 0.
    \item Remember: adjacency plays a big role in the mapping and selection of 
    cells, whether it's physical (horizontal/vertical) adjacency or logical 
    (wrap around) adjacency.
\end{itemize}
