\section{Introduction}

\subsection{Boolean Algebra}
Boolean Algebra provides the mathematical operations and rules for working with the set $\{0, 1\}$, serving as the foundation for digital logic and modern computer design. As detailed in Kenneth Rosen’s \textit{Discrete Mathematics and Its Applications}, this system was first introduced by George Boole in his 1854 work, \textit{The Laws of Thought}. While it differs from elementary algebra by dealing with logical values rather than real numbers, its practical power was realized in 1938 when Claude Shannon demonstrated its use in designing switching circuits.

\subsection{Boolean Expressions}
\textbf{Boolean Variables} \\
A Boolean variable is a fundamental data type in Computer Science that stores one of two logical values: True (1) or False (0). These variables take values only from the set $B = \{0,1\}$.


\noindent \textbf{Boolean Expressions} \\
These are formed using variables and the three primary operators:
\begin{itemize}
    \item \textbf{Complementation (NOT):} $\bar{x}$ (reverses the bit).
    \item \textbf{Boolean Sum (OR):} $x + y$ (is 1 if $x$ or $y$ is 1).
    \item \textbf{Boolean Product (AND):} $x \cdot y$ (is 1 only if both are 1).
\end{itemize}


\subsection{Boolean Functions}
A function $f: B^n \to B$ that assigns a binary value to each $n$-tuple of binary variables. These are usually represented using \textbf{Truth Tables}.

\subsection{Identities Of Boolean Algebra}
These are fundamental laws that define how binary variables \textbf{0} and \textbf{1} are manipulated, simplifying logical expressions and digital circuits.

\begin{itemize}
    \item \textbf{Identity Law ($x \cdot 1 = x$ and $x + 0 = x$)} \\
    Multiplying by 1 or adding 0 does not change the variable. \\
    Example: If $x = 0$, then $0 \cdot 1 = 0$ and $0 + 0 = 0$.
    
    \item \textbf{Domination Law ($x + 1 = 1$ and $x \cdot 0 = 0$)} \\
    Adding 1 dominates the expression (forcing it to 1), and multiplying by 0 dominates it (forcing it to 0). \\
    Example: If $x = 0$, then $0 + 1 = 1$ and $0 \cdot 0 = 0$.

    \item \textbf{Idempotent Law ($x + x = x$ and $x \cdot x = x$)} \\
    Combining a variable with itself (using AND or OR) results in the variable itself. There are no powers or multiples in Boolean algebra. \\
    Example: If $x = 1$, then $1 + 1 = 1$ (there is no 2 in Boolean algebra), $1 \cdot 1 = 1$.
    
    \item \textbf{Double Complement ($\overline{\overline{x}} = x$)} \\
    Taking the NOT of a NOT cancels out. \\
    Example: If $x = 1$, then $\bar{1} = 0$, and the complement of that is $\bar{0} = 1$.
    
    \item \textbf{Commutative Law ($x + y = y + x$ and $x \cdot y = y \cdot x$)} \\
    The order of the variables does not matter. \\
    Example: $1 + 0 = 0 + 1$. Both result in 1.
    
    \item \textbf{Associative Law ($(x + y) + z = x + (y + z)$)} \\
    When using the same operator, the grouping of variables does not change the result. \\
    Example: $(1 + 0) + 1 = 1 + (0 + 1) = 1$.
    
    \item \textbf{Distributive Law ($x \cdot (y + z) = (x \cdot y) + (x \cdot z)$)} \\
    Just like in regular math, you can distribute the product over the sum. Boolean algebra also allows distributing the sum over the product. \\
    Example: $1 \cdot (0 + 1) = (1 \cdot 0) + (1 \cdot 1) \implies 1 \cdot 1 = 0 + 1 \implies 1 = 1$.
    
    \item \textbf{De Morgan’s Law ($\overline{x \cdot y} = \bar{x} + \bar{y}$ and $\overline{x + y} = \bar{x} \cdot \bar{y}$)} \\
    The complement of a product is the sum of the complements, and the complement of a sum is the product of the complements. \\
    Example: $\overline{1 \cdot 0} = \bar{1} + \bar{0} \implies \bar{0} = 0 + 1 \implies 1 = 1$.
    
    \item \textbf{Absorption Law ($x + (x \cdot y) = x$ and $x \cdot (x + y) = x$)} \\
    The smaller term $(x \cdot y)$ is absorbed by the $x$. If $x$ is true, the whole thing is true regardless of $y$, same applies for $(x + y)$. \\
    Example: $1 + (1 \cdot 0) = 1 + 0 = 1$. The $y$ variable is absorbed.
\end{itemize}