\section{Modern Applications of Boolean Algebra}
Besides circuit design, Boolean algebra remains essential in cryptography, biological network modeling, and information retrieval, where binary logic is used for efficient decision making, secure computation, and structured search .
\cite{picek2016cryptographic}

\subsection{Cryptography}
One of the most important modern uses of Boolean algebra is in cryptography. Boolean functions are building blocks of symmetric cryptograhic systems, and researchers study their properties such as nonlinearity, algebraic degree, and resistance to attacks in order to design stronger ciphers. \cite{tu2012boolean}

\subsection{Biological Network Modeling}
Boolean algebra is also used in bioinformatics and systems biology to model gene regulatory networks. In these models, genes are represented as binary states, such as active or inactive, which makes Boolean logic a natural way to describe complex biological interactions. Recent research shows that Boolean networks are useful for reconstructing regulatory topology from gene-expression data and for understanding how multiple regulators jointly control target genes.\cite{puvsnik2022review}

\subsection{Information Retieval and Search}
Boolean algebra is widely used in information retrieval, especially in query formulation and filtering. Search systems use AND, OR, and NOT logic to combine keywords, narrow results, and improve precision in document retrieval. Research on Boolean query refinement also shows that logical query transformations can improve recall and precision in systematic review searching, which is important in evidence based research. \cite{alharbi2020refining}

\subsection{Database and Data Filtering}
Boolean logic is central to database querying and filtering because it helps systems return records that satisfy specific conditions. In practical terms, Boolean expressions are used in selection rules, access filtering, and indexing strategies to improve speed and accuracy when working with large datasets. This makes Boolean algebra important not only for hardware design but also for software systems that manage and retrieve information efficiently.