\section{Quine-McCluskey Method}

\subsection{Definition}
The Quine-McCluskey method is a tabular, algorithmic approach to Boolean
function minimisation. It is essentially a more systematic and
computer friendly version of the Karnaugh map method. This method is a
deterministic algorithm that can be programmed and is effective for any
number of variables.

\subsubsection{How It Works}

The Quine-McCluskey method proceeds in two well-defined stages.

\begin{description}
    \item[Step 1: Finding prime implicants.]
    This step identifies the largest possible groups of 1s and don't cares
    that can be combined. It works by first listing all minterms and don't
    cares in binary and sorting them into groups according to the number of
    1-bits in their binary representation. Terms from adjacent groups are
    then compared and merged if they differ in exactly one bit
    position, using the algebraic identity:
    \[
        AB + AB' = A
    \]
    The differing bit is replaced with a dash ($-$) to indicate elimination.
    This process repeats on the resulting terms until no further merging is
    possible. Any term that cannot be merged further is a prime implicant.

    \item[Step 2: Finding the essential prime implicants.]
    This step selects the minimal set of prime implicants needed to cover
    all the original minterms, the 1s of the function. A
    prime implicant chart is constructed with prime implicants as
    rows and original minterms as columns. A prime implicant is
    essential if it is the only one covering a particular minterm, it must be included in the final expression. After selecting all
    essential prime implicants, any uncovered minterms are resolved by
    choosing additional prime implicants, preferring those that cover the
    most remaining minterms. This is a classic set-covering problem.
\end{description}

\subsubsection{Example}

Given: \(f(A,B,C,D) = \sum m(0,1,2,5,6,7,8,9,10,14)\)

All minterms are listed in binary and sorted by their number of 1 bits,
as shown in Table~\ref{tab:grouping}. Adjacent groups are then compared and
combined where possible, producing the first-round combinations in
Table~\ref{tab:combinations}.Combining continues on these results until no further merging is possible.
Terms that remain uncombined are the prime implicants. These are entered
into the prime implicant chart, essential prime implicants are identified,
and the minimised Sum of Products expression is written.

\begin{table}[h!]
\centering
\caption{Minterms grouped by number of 1-bits}
\label{tab:grouping}
\begin{tabular}{@{}ccc@{}}
\toprule
\textbf{\# of 1s} & \textbf{Minterm} & \textbf{A\;B\;C\;D} \\
\midrule
0 & 0  & 0\;0\;0\;0 \\
\midrule
1 & 1  & 0\;0\;0\;1 \\
  & 2  & 0\;0\;1\;0 \\
  & 8  & 1\;0\;0\;0 \\
\midrule
2 & 5  & 0\;1\;0\;1 \\
  & 6  & 0\;1\;1\;0 \\
  & 9  & 1\;0\;0\;1 \\
  & 10 & 1\;0\;1\;0 \\
\midrule
3 & 7  & 0\;1\;1\;1 \\
  & 14 & 1\;1\;1\;0 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h!]
\centering
\caption{First-round combinations}
\label{tab:combinations}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Minterms} & \textbf{Result} & \textbf{A\;B\;C\;D} \\
\midrule
0,\;1   & (0,1)  & 0\;0\;0\;-- \\
0,\;2   & (0,2)  & 0\;0\;--\;0 \\
0,\;8   & (0,8)  & --\;0\;0\;0 \\
1,\;9   & (1,9)  & --\;0\;0\;1 \\
2,\;6   & (2,6)  & 0\;--\;1\;0 \\
2,\;10  & (2,10) & --\;0\;1\;0 \\
8,\;9   & (8,9)  & 1\;0\;0\;-- \\
8,\;10  & (8,10) & 1\;0\;--\;0 \\
6,\;7   & (6,7)  & 0\;1\;1\;-- \\
6,\;14  & (6,14) & --\;1\;1\;0 \\
\bottomrule
\end{tabular}
\end{table}
